Reasoning about a Capability Machine with Local Capabilities\\ Provably Safe Stack and Return Pointer Management (without OS Support)

### 2017-02-15 csf-2017/paper.tex
Shorten.

The final semantics we discuss is that of $\store{}{}$. As expected, it updates
the memory to the requested value if the capability has write authority for the
requested location. However, the instruction is complicated by the presence of
\emph{local capabilities}, modeled after the ones in
the CHERI processor~\cite{Watson2015Cheri}. Basically, local capabilities are special in that
they can only be kept in registers, i.e.\ they cannot be stored to memory. This
means that local capabilities can be \emph{temporarily} given to an adversary,
for the duration of an invocation: if we take care to erase the capability from
the register file after control is passed back to us, we can be sure that they
will not have been able to store the capability. However, there is one exception
to the rule above: local capabilities can be stored to a part of memory for
which we have a capability with write-local authority (i.e.\ permission $\rwl$ or
$\rwlx$). This exception is intended to accomodate regions of memory that behave
like the C stack: stashing away the contents of registers on the stack can be
allowed, even if this includes local capabilities. As long as all capabilities
with write-local authority are themselves local and the stack is cleared after
control is passed back by the adversary, we will see that this does not break
the intended behavior of local capabilities.

We point out that our local capabilities capture only a part of the semantics of
local capabilities in CHERI. Specifically, on top of the above restrictions,
CHERI's default implementation of CCall additionally forbids local capabilities
from being passed accross module boundaries, something which we rely on
crucially in this paper, for passing around local return pointers and stack
capabilities. However, CHERI's CCall is not implemented in hardware, but in
software, precisely to allow experimenting with alternative models like ours.

@@

In order to have a reasonably realistic system, we have created a
simple model of linking, so programs can have access to other programs
through a linking table. We also assume malloc to be part of the
trusted computing base satisfying a certain specification. Malloc and
the model linking is described to the degree necessary, in the next
section, but we refer to the technical appendix~\cite{technical_appendix} for further
details.

\section{Stack and return pointer management using local capabilities}
\label{sec:stack-and-return-pointer}
%\lau{Do we want to use the term ``pointer'' at any point? I have also used it, but I guess it might create some confusing to whether we have traditional pointer as well as capabilities?}
One of the contributions in this paper is a demonstration that local
capabilities on a capability machine can be used in a calling convention that
enforces control-flow correctness in a way that is provably watertight,
potentially efficient, does not rely on a trusted central stack manager, and
with support for exposing higher-order interfaces to an adversary. In this
section, we explain this convention. We start by explaining the high-level
approach, the different issues that need to be taken into account, starting with
first-order interfaces and moving onto a higher-order setting next. After that,
we define a number of reusable macro-instructions that can be used to
conveniently apply the proposed convention in subsequent examples.

The basic idea of our approach is quite simple: we stick to a single, rather
standard, C stack and register-passed return pointers, much like a standard C
ABI. However, to prevent various ways of
misusing this basic scheme, we put local capabilities to work and take a number
of not-always-obvious safety measures. In the next paragraphs, we will explain
the issues to be considered in all the relevant situations: when (1) starting
our program, (2) Returning to the adversary, (3) invoking the adversary, (4)
returning from the adversary, (5) invoking an adversary callback and (6) having
a callback invoked by the adversary.

% \dominique{Add a summary table somewhere?}
% \lau{It seems like we want have the space for this}.

\textbf{Program startup}
When our code starts executing, we assume that the language runtime initializes
the memory as follows: a contiguous array of memory is reserved for the stack,
for which we receive a stack pointer in a register, $r_\stk$, that we reserve for this
purpose. We stress that the stack is not built-in, but merely an
abstraction we put on this piece of the memorey. This stack pointer must be local and it must carry $\rwlx$ permission. It is crucial for
our approach that the stack is the only part of memory for which there will ever
be an obtainable capabilities with $\rwlx$ or $\rwl$ capabilities. We expect the language
runtime to enforce this by not handing out such capabilities for other parts of
memory in any situation (i.e.\ not through malloc, not during loading or linking
etc.). Additionally, in our examples, we typically also assume some static parts
of memory where instructions are stored or static data. Another part of memory
(called the heap) is initially governed by malloc and at program startup, no
other code has capabilities for this memory. Malloc will hand out capabilities
for allocated regions as requested, but it will never hand out capabilities with
write-local capability (i.e.\ no $\rwlx$ or $\rwl$ permissions). For simplicity,
we assume that memory allocated through malloc cannot be freed.

\textbf{Returning to the adversary} Perhaps the simplest scenario is the case of
returning to the adversary after they invoked our code. In this case, we will
have received a return pointer from them, and we just need to jump to it as
usual. The only thing we need to take care of is to properly clear the registers
before we do (erasing any data we do not want to pass as (part of) the return
value) and similarly. However, it is important to carefully consider what needs
to happen with the stack. Since we may have used the stack for various purposes
(register spilling, storing local state when invoking other functions etc.), we
clearly need to erase that data before returning to the adversary, to prevent
the adversary access to it.

However, if we are returning from a function that has itself invoked adversary
code, then even the unused part of the stack may still contain data and
capabilities left there by them, including local capabilities since the stack is
write-local. The adversary could take advantage of this by using the unused part
of the stack to store either stack pointers or return pointers from such
previous invocations and then use them when they're no longer supposed to. To
prevent this, before returning, we need to clear (i.e.\ overwrite with zeros)
the entire part of the stack that the adversary has had access to, not just the
parts that we have used ourselves. Essentially, this is needed for the same
reason that we clear the register file: because the adversary should not be able
to keep hold of local capabilities when control is passed to the trusted code
and back to the adversary. Clearing the entire unused part of the stack is the
most problematic aspect of our calling convention in terms of efficiency, since
it may consist of a large part of memory (but see Section~\ref{sec:discussion}
for how this might be mitigated).

\textbf{Invoking the adversary} Slightly more complex is the case where we
invoke the adversary. As above, we clear all the registers, except those that
contain arguments the adversary is intended to have access to. We leave a copy
of the stack pointer in $r_\stk$, but only after we have used
the $\subseg{}{}{}$ to shrink its authority to just the part of the stack that
we are not using. In one of the registers, we also provide a return pointer,
which we will say more about below.

Already important is the fact that the return pointer must be provided to the
adversary as a local capability. If not, the adversary would be able to store
away the return pointer in a global data structure, and they would be able to jump
to it in circumstances where this should not be possible: for example, they may
store the return pointer, legally jump to it a first time, wait to be invoked
again and then jump to the old return pointer a second time, rather than the new
return pointer received for the new invocation. Similarly, they may store the
return pointer, invoke a function in our code, wait for us to invoke him again
and then jump to the stored, old return pointer rather than the new one they
received upon the second invocation.
% \lau{If we need space, then maybe one of the examples is sufficient.}
% \lau{Both of these are good to have their as the two last examples of Section~\ref{sec:examples} use them.}
%
% \lau{In the examples, we refer to the local return pointer as
%   ``protected return pointer''. We should either introduce that here
%   or use a different formulation in the examples. }
% \lau{It seemed did not seem to add anything to adopt the terminlogy - we
%   often highlight that the return pointer is local which in most cases
%   is so important that we do not want to leave it out.}

@@

A final required safety measure is that before invocation of the adversary, we
need to clear the part of the stack that we are not using. Otherwise, similarly
to what we explained above for returning to the adversary, that part of the
stack may still contain local capabilities from previously executed code and the
adversary could exploit this in much the same way as above.

@@

\begin{figure}[htbp]
  \centering
  \begin{align*}
  & \readCond{}(\gl)(W) =  \\
  & \left\{ \npair{(\start,\addrend)} \; \middle| \;
    \begin{aligned}
      & \exists r \in \var{localityReg}(g,W) \ldotp \\
      & \;\; \exists [\start',\addrend'] \supseteq [\start,\addrend] \ldotp W(r)\nsubsim[n] \iota_{\start',\addrend'}^\pwl 
    \end{aligned} \right\}\\
  & \writeCond{}(\iota,\gl)(W) =  \\
  & \left\{
    \npair{(\start,\addrend)}
    \; \middle| \;
    \begin{aligned}
      & \exists r \in \var{localityReg}(g,W) \ldotp \\
      & \;\; W(r) \text{\footnotesize{ is address-stratified and}} \\
      & \;\; \exists [\start',\addrend'] \supseteq [\start,\addrend] \ldotp W(r)\nsupsim[n-1] \iota_{\start',\addrend'}
    \end{aligned} \right\}\\
  & \execCond{}(\gl)(W) = \\
  & \left\{
    \begin{multlined}
      \npairP{(\var{perms}, \\
        \start,\addrend)}
    \end{multlined}
     \; \middle| \;
    \begin{aligned}
      & \forall n' < n, W' \future W, a \in [\start,\addrend], \perm \in \var{perms} \ldotp \\
      & \;\;\;\; \npair[n']{((\perm,\gl),\start,\addrend,\addr)} \in \stder(W')\\
    \end{aligned} \right\} \\
  & \entryCond{}(\gl)(W) = \\
  & \left\{ \npair{(\start,\addrend,\addr)} \; \middle| \;
    \begin{aligned}
 &  \forall n' < n \ldotp \forall W' \future W \ldotp\\
      & \quad \npair[n']{((\exec,\gl),\start,\addrend,\addr)} \in \stder(W')
    \end{aligned} \right\} \\
  & \quad \text{where } \gl = \local \Rightarrow \future = \futurewk \\
  & \quad \text{and } \gl = \glob \Rightarrow \future = \futurestr
  \end{align*}
\caption{Permission-based conditions}
\label{fig:perm-conds}
\end{figure}
}
%subsection introduction
% \dominique{I would try to motivate the LR defs by saying that the LR is
%   constructed to express the FTLR and the FTLR says that if a piece of code only
%   has access to values respecting the invariants in a world, then it will
%   respect them itself too.} 
The logical relation defines semantically what it
means to be capability safe.
% \dominique{We have not introduced capability
%   safety..} 
In large parts, capability safety is defined by when a capability
can be safely given to an adveserial piece of code. In the following paragraphs,
we will go through the definitions of the logical relation and provide an
intuition why they define capability safety. The definition of the logical
relation is found in \figurename~\ref{fig:logrel}, and \figurename~\ref{fig:perm-conds}.
Some definitions are omitted and explained when needed. The precise definitions
can be found in the technical appendix~\cite{technical_appendix}.

@@

\subsection{Safety of the capability machine}
With the logical relation defined, we can now state the fundamental theorem of
our logical relation: a strong theorem that formalises the guarantees offered by
the capability machine. Essentially, it says that a capability that only grants
safe authority is capability safe as a program counter.
\begin{theorem}[Fundamental Theorem]
  \label{thm:ftlr}
  If one of the following holds:
  \begin{align*}
      & \bullet
        \begin{aligned}[t]
        &\perm = \exec \text{\footnotesize{ and }}\npair{(\start,\addrend)} \in \readCond{}(\gl)(W)
      \end{aligned} \\
    & \bullet 
      \begin{aligned}[t]
        &\perm = \rwx \text{\footnotesize{ and }} \\
        &\npair{(\start,\addrend)} \in \readCond{}(\gl)(W) \text{\footnotesize{ and }}\\
        &\npair{(\start,\addrend)} \in \writeCond{}(\iota^\nwl,\gl)(W)
      \end{aligned} \\
    & \bullet 
      \begin{aligned}[t]
        &\perm = \rwlx \text{\footnotesize{ and }}\\
        &\npair{(\start,\addrend)} \in \readCond{}(\gl)(W) \text{\footnotesize{ and }}\\
        &\npair{(\start,\addrend)} \in \writeCond{}(\iota^\pwl,\gl)(W),
      \end{aligned}
  \end{align*}
  then $\npair{((\perm,\gl),\start,\addrend,\addr)} \in \stder(W)$
\end{theorem}
The Fundamental Theorem allows us to prove capability safety for
adveserial programs: if the authority granted by
the capabilty for the untrustred program is reasonable, then it is safe to
execute. Intuitively, this is because our logical relation captures the
safe behaviours of the capability machine that do not allow you to
produce something \emph{unsafe} from something safe. 
The Fundamental Theorem also serves as a sanity check of the
logical relation as it guarantees that a large class of capabilities
actually inhabits it.

@@

\subsection{Encapsulation of local state}
The capability machine provides encapsulation of local state. The programs
\texttt{\footnotesize{f1}} and \texttt{\footnotesize{f2}} in
\figurename~\ref{fig:prog-f1-and-f2} demonstrate this. They are very similar: both
store some local state, call an untrusted piece of code ($\adv$), and then test
whether the local state is unchanged. They differ in the way they do this.
Program \texttt{\footnotesize{f1}} assumes it has access to a stack pointer and uses the stack
to store its local state. It also uses the stack-based calling convention
\texttt{\footnotesize{scall}} to call the adversary. On the other hand, \texttt{\footnotesize{f2}} uses
malloc to allocate memory for its local state and uses an activation-record
based calling convention (described in the technical appendix) to run the
adversarial code.

@@

\section{Discussion}
\label{sec:discussion}
\subsection{Calling convention}
\textbf{Formulating control flow correctness} While we claim that our calling convention enforces control-flow
correctness, we do not prove a general theorem that shows this, it is not clear what such a theorem should state. Formulations in terms of a control-flow graph, like the one by
\cite{abadi_control-flow_2005}, do not take into account temporal properties,
like the well-bracketedness that Example~\texttt{\footnotesize{g1}} relies on.
In fact, our examples show that our logical relation imply a form of
control-flow correctness that is stronger than such formulations, although this
is not made very explicit. As future work, we consider looking at a more
explicit and useful way to formalise well-bracketed control flow. The idea would
be to define a variant of our capability machine with built-in call and return
instructions and well-bracketed control flow built-in to the operational
semantics, and then prove that compiling such code to our machine using our
calling convention is fully abstract~\cite{abadi_protection_1998}, which roughly
means that the lower-level machine does not allow interactions with the compiled
code that do not correspond to an interaction with the original code in the
high-level machine.

\textbf{Performance and the requirement for stack clearing} 
The additional security measures of the calling convention described in
Section~\ref{sec:stack-and-return-pointer} impose an overhead compared to
a calling convention that does not provide security. However, the security measures generally consist of a
few checks or register clearings on every boundary crossing between trusted code
and adversary, which should impose an acceptable performance overhead. The only
exception is the requirements for stack clearing which we occur in two
situations: when returning to the adversary and when invoking an adversary
callback. As we have explained, it is not sufficient to clear just the part of
the stack that we have used ourselves, but we need to clear all of the stack that we are not using ourselves. 
In other words on every boundary cross between trusted code and adversary code, a potentially large region of memory must be erased.
We believe this is a common requirement for typical usage scenarios of local capabilities and
capability machines like CHERI should consider to provide special support for
this requirement, in the form of a highly-optimized instruction for erasing
a large block of memory. We have no experience in processor design,
but we imagine this might be optimized as a vector instruction or even further,
by having the processor keep track of a range of memory for which loads should
pretend that it has already been cleared, by returning zero irrespective of the
actual contents of the memory.

\textbf{Modularity} It is important that our calling convention is modular.
This means that the requirements we have on callbacks, return
pointers etc. provided to trusted code by an adversary are also satisfied by
callbacks, return pointers etc. that we construct ourselves. For example, our
return pointers are local capabilities because they must point to
memory where we can store the old stack pointer, but the adversary's return
pointers are also allowed to be local. Adversary callbacks on the other hand are
required to be global but the callbacks we construct ourselves are also
allocated on the heap and global. In other words, we do not assume that our code
has some sort of special status and it is possible for the adversary to protect
themselves against us the same way we protect against them.

@@

\textbf{Public/private future worlds} A novel aspect of our logical relation is
how we model the temporary nature of local capabilities using public/private
future worlds. The main insight is that the temporary, revokable nature of these
capabilities can be seen as a generalisation of the syntactically-enforced
unstoreable status of evaluation contexts in lambda calculi without control
effects (of which well-bracketed control flow is a consequence). To reason about
code that relies on this (particularly, the original awkward example),
\citet{Dreyer:jfp12} (DNB) formally capture the special status of evaluation
contexts using Kripke worlds with public and private future world relations.
Essentially, they allow relatedness of evaluation contexts to be monotone with
respect to a weaker future world relation (public) than relatedness of values,
formalising the idea that it is safe to make temporary internal state
modifications (private world transititions, that invalidate the continuation,
but not other values) while an expression is performing internal steps, as long
as the code returns to a stable state (i.e. transitions to a public future world
of the original) before returning (i.e. invoking the continuation). We
generalise this idea to reason about local capabilities: validity of local
capabilities is allowed to be monotone with respect to a weaker future-world
relation than other values, which we can exploit to distinguish between state
changes that are always safe (public future worlds) and changes that are only
valid if we clear all local capabilities (private future worlds). Our future
world relations are similar to \citet{Dreyer:jfp12}'s (for example, our proof of
the awkward example uses exactly the same state transition system),
but they turn up in an entirely different place in the logical relation: rather
than using public future worlds for the special syntactic category of evaluation
contexts, they are used in the value relation depending on the
locality of the capability at hand. Additionally, our worlds are a bit more
complex because, to allow local memory capabilities and write-local
capabilities, they can contain (revokable) temporary regions that are only
monotonous w.r.t. public future worlds, while DNB's worlds are entirely
permanent.

\textbf{Local capabilities in high-level languages} It is worth mentioning that
local capabilities are quite similar to a feature proposed for the high-level
language Scala: \citet{osvald_gentrification_2016}'s second-class or local
values. They are a kind of values that can be provided to other code for
immediate use without allowing it to be stored in a closure or reference for
later use. We believe reasoning about such values will require techniques similar 
to the ones we have used for local capabilities.

>>>>

Finally, the $\store{}{}$ instruction updates the memory to the requested value
if the capability has write authority for the requested location. However, the
instruction is complicated by the presence of \emph{local capabilities}, modeled
after the ones in the CHERI processor~\cite{Watson2015Cheri}. Basically, local
capabilities are special in that they can only be kept in registers, i.e.\ they
cannot be stored to memory. This means that local capabilities can be
\emph{temporarily} given to an adversary, for the duration of an invocation: if
we take care to erase the capability from the register file after control is
passed back to us, they will not have been able to store the capability.
However, there is one exception to the rule above: local capabilities can be
stored to memory for which we have a capability with write-local authority
(i.e.\ permission $\rwl$ or $\rwlx$). This is intended to accomodate a stack,
where register contents can be stored, including local capabilities. As long as all
capabilities with write-local authority are themselves local and the stack is
cleared after control is passed back by the adversary, we will see that this
does not break the intended behavior of local capabilities.

We point out that our local capabilities capture only a part of the semantics of
local capabilities in CHERI. Specifically, in addition to the above, CHERI's
default implementation of the CCall exception handler forbids local capabilities
from being passed accross module boundaries. Such a restriction fundamentally
breaks our calling convention, since we pass around local return pointers and
stack capabilities. However, CHERI's CCall is not implemented in hardware, but
in software, precisely to allow experimenting with alternative models like ours.

@@

In order to have a reasonably realistic system, we use a simple model of
linking, so programs can have access to other programs through a linking table.
We also assume malloc to be part of the trusted computing base satisfying a
certain specification. Malloc and linking tables are described further in the
next section, but we refer to the technical appendix~\cite{technical_appendix}
for full details.

\section{Stack and return pointer management using local capabilities}
\label{sec:stack-and-return-pointer}
One of the contributions in this paper is a demonstration that local
capabilities on a capability machine support a calling convention that enforces
control-flow correctness in a way that is provably watertight, potentially
efficient, does not rely on a trusted central stack manager, and supports
higher-order interfaces to an adversary. In this section, we explain this
convention's high-level approach, the security measures to be taken in a number
of situations (motivating each separately with a summary table at the end).
After that, we define a number of reusable macro-instructions that can be used
to conveniently apply the proposed convention in subsequent examples.

The basic idea of our approach is simple: we stick to a single, rather standard,
C stack and register-passed return pointers, much like a standard C calling
convention. However, to prevent various ways of misusing this basic scheme, we
put local capabilities to work and take a number of not-always-obvious safety
measures. In the next paragraphs, we will explain the issues to be considered in
all the relevant situations: when (1) starting our program, (2) returning to the
adversary, (3) invoking the adversary, (4) returning from the adversary, (5)
invoking an adversary callback and (6) having a callback invoked by the
adversary.

\textbf{Program startup} We assume that the language runtime initializes the
memory as follows: a contiguous array of memory is reserved for the stack, for
which we receive a stack pointer in a special register $r_\stk$. We stress that
the stack is not built-in, but merely an abstraction we put on this piece of the
memory. The stack pointer is local and has $\rwlx$ permission. Crucially, the
stack is the only part of memory for which the runtime (including malloc,
loading, linking) will ever provide $\rwlx$ or $\rwl$ capabilities.
Additionally, our examples typically also assume some memory to store
instructions or static data. Another part of memory (called the heap) is
initially governed by malloc and at program startup, no other code has
capabilities for this memory. Malloc hands out $\rwx$ capabilities for allocated
regions as requested (no $\rwlx$ or $\rwl$ permissions). For simplicity, we
assume that memory allocated through malloc cannot be freed.

\textbf{Returning to the adversary} Perhaps the simplest situation is returning
to the adversary after they invoked our code. In this case, we have received a
return pointer from them, and we just need to jump to it as usual. An obvious
security measure to take care of is properly clearing the non-return-value
registers before we jump (since they may contain data or capabilities that the
adversary should not get access to). Additionally, we may have used the stack
for various purposes (register spilling, storing local state when invoking other
functions etc.), so we also need to erase that data before returning to the
adversary.

However, if we are returning from a function that has itself invoked adversary
code, then clearing the used part of the stack is not enough. The \emph{unused}
part of the stack may also contain data and capabilities, left there by the
adversary, including local capabilities since the stack is store-local. The
adversary could take advantage of this by using the unused part of the stack to
store either stack pointers or return pointers from such previous invocations
and then use them when they're no longer supposed to. To prevent this, we need
to clear (i.e.\ overwrite with zeros) the entire part of the stack that the
adversary has had access to, not just the parts that we have used ourselves.
This requirement may seem strong, but it is essentially there for the same
reason that we clear the register file: the adversary should not be able to keep
hold of local capabilities when control is passed to trusted code and back to
the adversary. Since we may be talking about a large part of memory, this is the
most problematic aspect of our calling convention for performance, but see
Section~\ref{sec:discussion} for how this might be mitigated.

\textbf{Invoking the adversary} A slightly more complex case is invoking the
adversary. As above, we clear all the non-argument registers, as well as the
part of the stack that we are not using (because, as above, it may contain local
capabilities from previously executed code that the adversary could exploit in
the same way as above). We leave a copy of the stack pointer in $r_\stk$, but
only after we have used the $\subseg{}{}{}$ to shrink its authority to the part
that we are not using ourselves.

In one of the registers, we also provide a return pointer, which must be a local
capability. If it were global, the adversary would be able to store away the
return pointer in a global data structure, and jump to it later, in
circumstances where this should not be possible. For example, they could store
the return pointer, legally jump to it a first time, wait to be invoked again
and then jump to the old return pointer a second time, instead of the new return
pointer received for the second invocation. Similarly, they could store the
return pointer, invoke a function in our code, wait for us to invoke them again
and then jump to the old return pointer rather than the new one, received for the
second invocation.
% \lau{If we need space, then maybe one of the examples is sufficient.}
% \lau{Both of these are good to have their as the two last examples of Section~\ref{sec:examples} use them.}
%
% \lau{In the examples, we refer to the local return pointer as
%   ``protected return pointer''. We should either introduce that here
%   or use a different formulation in the examples. }
% \lau{It seemed did not seem to add anything to adopt the terminlogy - we
%   often highlight that the return pointer is local which in most cases
%   is so important that we do not want to leave it out.}

@@

@@

\begin{figure}[htbp]
  \centering
  \begin{align*}
  & \readCond{}(\gl)(W) =  \\
  & \left\{ \npair{(\start,\addrend)} \; \middle| \;
    \begin{aligned}
      & \exists r \in \var{localityReg}(g,W) \ldotp \\
      & \;\; \exists [\start',\addrend'] \supseteq [\start,\addrend] \ldotp W(r)\nsubsim[n] \iota_{\start',\addrend'}^\pwl 
    \end{aligned} \right\}\\
  & \writeCond{}(\iota,\gl)(W) =  \\
  & \left\{
    \npair{(\start,\addrend)}
    \; \middle| \;
    \begin{aligned}
      & \exists r \in \var{localityReg}(g,W) \ldotp \\
      & \;\; W(r) \text{\footnotesize{ is address-stratified and}} \\
      & \;\; \exists [\start',\addrend'] \supseteq [\start,\addrend] \ldotp W(r)\nsupsim[n-1] \iota_{\start',\addrend'}
    \end{aligned} \right\}\\
  & \execCond{}(\gl)(W) = \\
  & \left\{
    \begin{multlined}
      \npairP{(\var{perms}, \\
        \start,\addrend)}
    \end{multlined}
     \; \middle| \;
    \begin{aligned}
      & \forall n' < n, W' \future W, a \in [\start,\addrend], \perm \in \var{perms} \ldotp \\
      & \;\;\;\; \npair[n']{((\perm,\gl),\start,\addrend,\addr)} \in \stder(W')\\
    \end{aligned} \right\} \\
  & \entryCond{}(\gl)(W) = \\
  & \left\{ \npair{(\start,\addrend,\addr)} \; \middle| \;
    \begin{aligned}
 &  \forall n' < n \ldotp \forall W' \future W \ldotp\\
      & \quad \npair[n']{((\exec,\gl),\start,\addrend,\addr)} \in \stder(W')
    \end{aligned} \right\} \\
  & \quad \text{where } \gl = \local \Rightarrow \future = \futurewk \\
  & \quad \text{and } \gl = \glob \Rightarrow \future = \futurestr
  \end{align*}
\caption{Permission-based conditions}
\label{fig:perm-conds}
\end{figure}
}
%subsection introduction
% \dominique{I would try to motivate the LR defs by saying that the LR is
%   constructed to express the FTLR and the FTLR says that if a piece of code only
%   has access to values respecting the invariants in a world, then it will
%   respect them itself too.} 
The logical relation defines semantically what it
means to be capability safe.
% \dominique{We have not introduced capability
%   safety..} 
In large parts, capability safety is defined by when a capability
can be safely given to an untrusted piece of code. In the following paragraphs,
we will go through the definitions of the logical relation and provide an
intuition why they define capability safety. The definition of the logical
relation is found in \figurename~\ref{fig:logrel}, and \figurename~\ref{fig:perm-conds}.
Some definitions are omitted and explained when needed. The precise definitions
can be found in the technical appendix~\cite{technical_appendix}.

@@

\subsection{Safety of the capability machine}
With the logical relation defined, we can now state the fundamental theorem of
our logical relation: a strong theorem that formalises the guarantees offered by
the capability machine. Essentially, it says that a capability that only grants
safe authority is capability safe as a program counter.
\begin{theorem}[Fundamental Theorem]
  \label{thm:ftlr}
  If one of the following holds:
  \begin{align*}
      & \bullet
        \begin{aligned}[t]
        &\perm = \exec \text{\footnotesize{ and }}\npair{(\start,\addrend)} \in \readCond{}(\gl)(W)
      \end{aligned} \\
    & \bullet 
      \begin{aligned}[t]
        &\perm = \rwx \text{\footnotesize{ and }} \\
        &\npair{(\start,\addrend)} \in \readCond{}(\gl)(W) \text{\footnotesize{ and }}\\
        &\npair{(\start,\addrend)} \in \writeCond{}(\iota^\nwl,\gl)(W)
      \end{aligned} \\
    & \bullet 
      \begin{aligned}[t]
        &\perm = \rwlx \text{\footnotesize{ and }}\\
        &\npair{(\start,\addrend)} \in \readCond{}(\gl)(W) \text{\footnotesize{ and }}\\
        &\npair{(\start,\addrend)} \in \writeCond{}(\iota^\pwl,\gl)(W),
      \end{aligned}
  \end{align*}
  then $\npair{((\perm,\gl),\start,\addrend,\addr)} \in \stder(W)$
\end{theorem}
The Fundamental Theorem allows us to prove capability safety for
untrusted programs: if the authority granted by
the capabilty for the untrustred program is reasonable, then it is safe to
execute. Intuitively, this is because our logical relation captures the
safe behaviours of the capability machine that do not allow you to
produce something \emph{unsafe} from something safe. 
The Fundamental Theorem also serves as a sanity check of the
logical relation as it guarantees that a large class of capabilities
actually inhabits it.

@@

\subsection{Encapsulation of local state}
The capability machine provides encapsulation of local state. The programs
\texttt{\footnotesize{f1}} and \texttt{\footnotesize{f2}} in
\figurename~\ref{fig:prog-f1-and-f2} demonstrate this. They are very similar: both
store some local state, call an untrusted piece of code ($\adv$), and then test
whether the local state is unchanged. They differ in the way they do this.
Program \texttt{\footnotesize{f1}} assumes it has access to a stack pointer and uses the stack
to store its local state. It also uses the stack-based calling convention
\texttt{\footnotesize{scall}} to call the untrusted code. On the other hand, \texttt{\footnotesize{f2}} uses
malloc to allocate memory for its local state and uses an activation-record
based calling convention (described in the technical appendix) to run the
adversarial code.

@@

\section{Discussion}
\label{sec:discussion}

\subsection{Calling convention}
\textbf{Formulating control flow correctness} The attentive reader may notice
that, while we claim that our calling convention enforces control-flow
correctness, we do not prove a general theorem to show that this is the case.
The reason for this is that it is not clear what such a theorem should look
like. Formulations in terms of a control-flow graph, like the one by
\cite{abadi_control-flow_2005}, do not take into account temporal properties,
like the well-bracketedness that Example~\texttt{\footnotesize{g1}} relies on.
In fact, our examples show that our logical relation imply a form of
control-flow correctness that is stronger than such formulations, although this
is not made very explicit. As future work, we are considering to look at a more
explicit and useful way to formalise well-bracketed control flow. The idea would
be to define a variant of our capability machine with built-in call and return
instructions and well-bracketed control flow built-in to the operational
semantics, and then prove that compiling such code to our machine using our
calling convention is fully abstract~\cite{abadi_protection_1998}, which roughly
means that the lower-level machine does not allow interactions with the compiled
code that do not correspond to an interaction with the original code in the
high-level machine.

\textbf{Performance and the requirement for stack clearing} It is important to
consider the performance of our proposed calling convention.
Generally, the additional security measures described in
Section~\ref{sec:stack-and-return-pointer} obviously impose an overhead w.r.t.
an approach that does not provide security. However, they generally consist of a
few checks or register clearings on every boundary crossing between trusted code
and adversary, which should impose an acceptable performance overhead. The only
exception is the requirements that we have for stack clearing, in two
situations: when returning to the adversary and when invoking an adversary
callback. As we have explained, it is not sufficient to clear just the part of
the stack that we have used ourselves, but we need to clear the entire stack,
except the part that we are using ourselves. This means that on every crossing
of the boundary between trusted and adversary code, a potentially large region
of memory needs to be zeroed out entirely. We believe that this in fact is a
common requirement for typical usage scenario's of local capabilities and
capability machines like CHERI should consider to provide special support for
this requirement, in the form of a highly-optimized instruction for zeroing out
a potentially large block of memory. We have no experience in processor design,
but we imagine this might be optimized as a vector instruction or even further,
by having the processor keep track of a range of memory for which loads should
pretend that it has already been cleared, by returning zero irrespective of the
actual contents of the memory.

\textbf{Modularity} An important aspect of our calling convention is that it is
modular. By that, we mean that the requirements we have on callbacks, return
pointers etc. provided to trusted code by an adversary are also satisfied by
callbacks, return pointers etc. that we construct ourselves. For example, our
return pointers are necessarily local capabilities because they must point to
memory where we can store the old stack pointer, but the adversary's return
pointers are also allowed to be local. Adversary callbacks on the other hand are
required to be global but the callbacks we construct ourselves are also
allocated on the heap and global. In other words, we do not assume that our code
has some sort of special status and it is possible for the adversary to protect
themselves against us in the same way as we protect against them.

@@

\textbf{Public/private future worlds} A novel aspect of our logical relation is
how we model the temporary nature of local capabilities using public/private
future worlds. The main insight is that the temporary, revokable nature of these
capabilities can be seen as a generalisation of the syntactically-enforced
unstoreable status of evaluation contexts in lambda calculi without control
effects (of which well-bracketed control flow is a consequence). To reason about
code that relies on this (particularly, the original awkward example),
\citet{Dreyer:jfp12} (DNB) formally capture the special status of evaluation
contexts using Kripke worlds with public and private future world relations.
Essentially, they allow relatedness of evaluation contexts to be monotone with
respect to a weaker future world relation (public) than relatedness of values,
formalising the idea that it is safe to make temporary internal state
modifications (private world transititions, that invalidate the continuation,
but not other values) while an expression is performing internal steps, as long
as the code returns to a stable state (i.e. transitions to a public future world
of the original) before returning (i.e. invoking the continuation). We
generalise this idea to reason about local capabilities: validity of local
capabilities is allowed to be monotone with respect to a weaker future-world
relation than other values, which we can exploit to distinguish between state
changes that are always safe (public future worlds) and changes that are only
valid if we clear all local capabilities (private future worlds). Our future
world relations are similar to \citet{Dreyer:jfp12}'s (for example, our proof of
the awkward example uses exactly the same state transition system than theirs),
but they turn up in an entirely different place in the logical relation: rather
than using public future worlds for the special syntactic category of evaluation
contexts, they are used in the value relation depending on the depending on the
locality of the capability at hand. Additionally, our worlds are a bit more
complex because, to allow local memory capabilities and store-local
capabilities, they can contain (revokable) temporary regions that are only
monotonous w.r.t. public future worlds, while DNB's worlds are entirely
permanent.

\textbf{Local capabilities in high-level languages} It is worth mentioning that
local capabilities are quite similar to a feature proposed for the high-level
language Scala: \citet{osvald_gentrification_2016}'s second-class or local
values. They are a kind of values that can be provided to other code for
immediate use without allowing it to be stored in a closure or reference for
later use. We believe reasoning about such values will require very similar
reasoning techniques than the ones we have used for local capabilities.
